mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-05 Draft2.Rmd
text

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-11-09"
abstract: >
  This R package translates R objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic description, this new feature
  is a futher step towards reproducible science. The package can be extended by
  user-defined hooks that allow to replace variable names and new R functions by
  a suitable symbolic representation. We provide two working examples that
  illustrate the use of the new package in a scientific article and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.britannica.com/animal/bilby
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

@@

## Arithmetic operators
The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division. Additionally, we have provided two possible ways to represent multiplication: With sign and by juxtaposition and two different kinds of division.

```{r, results="asis", echo=TRUE}
catmath(x + y)

catmath(x - y)

# Note wrong precedence of custom operator
catmath(mean(m) %+-% 1.96 * s/sqrt(N))

# This is what we actually want
catmath(mean(m) %+-% (1.96 * s/sqrt(N)))

# Without parentheses
catmath(mean(m) %+-% {1.96 * s/sqrt(N)})

@@

@@

catmath(x^(y + z))

# without parenthesis
mathout(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))

@@

## Absolute value

@@

```{r, echo=FALSE}
# Improve display
hook(quote(mu_A), quote(mu["A"]))
hook(quote(mu_B), quote(mu["B"]))
hook(quote(sigma_A), quote(sigma["A"]))
hook(quote(sigma_B), quote(sigma["B"]))
hook(quote(mu_M), quote(mu["M"]))
hook(quote(zn), quote(z[tau]))
hook(quote(M), quote(overline(X)))
```

>>>>

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-11-09"
abstract: >
  This R package translates R objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic description, this new feature
  is a futher step towards reproducible science. The package can be extended by
  user-defined hooks that allow to replace variable names and new R functions by
  a suitable symbolic representation. We provide two working examples that
  illustrate the use of the new package in a scientific article and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

@@

## Hooks
Before start listing all the possible operations with \pkg{mathml},  we mention here that \pkg{mathml} also allows the user to create hooks for specific functions. As an example:

```{r, echo=TRUE}
# Improve display
hook(quote(mu_A), quote(mu["A"]))
hook(quote(mu_B), quote(mu["B"]))
hook(quote(sigma_A), quote(sigma["A"]))
hook(quote(sigma_B), quote(sigma["B"]))
hook(quote(mu_M), quote(mu["M"]))
hook(quote(zn), quote(z[tau]))
hook(quote(M), quote(overline(X)))
```

The hooks reported here will be the one used in Section @ref[] to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance

## Arithmetic operators
The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division. For instance:

```{r, results="asis", echo=TRUE}
catmath(x + y - z)
```

In addition, with \pkg{mathml} it is also possible to combine `minus` and `plus` and get a new operator. However, in R\ the operator \(\pm\) is not present and this leads to some interesting questions. If we let R\ present the expression, this is what we get:

```{r, results='asis', echo=TRUE}
# Note wrong precedence of custom operator
catmath(mean(m) %+-% 1.96 * s/sqrt(N))
```

As can be seen, the new custom operator takes precedence over multiplication, as R\ automatically puts `mean(m) %+-% 1.96` in parentheses. Unfortunately, this is not consistent with the rules of mathematics. As the reader well knows, it is not possible in R\ to change the precedence of custom operators. This leaves two possible solutions. The first involves a mathematically unnecessary parenthesis around `(1.96 * s/sqrt(N))`. The second is more elegant and mathematically consistent: here `1.96 * s/sqrt(N)` is put in curly brackets and mathematical formality is respected. In both solutions R\ is able to produce correct output once the order of operations is made explicit.

```{r, results='asis', echo=TRUE}
# This is what we actually want
catmath(mean(m) %+-% (1.96 * s/sqrt(N)))

# Without parentheses
catmath(mean(m) %+-% {1.96 * s/sqrt(N)})
```

_The following can be inserted in a table_

```{r, results='asis', echo=TRUE}

@@

At first glance, converting operators may seem trivial enough. However, when you want to deal with, for example, exponentiation, things can get more complicated. We see here that if we want `x` to the power of `y + z`, following R\ notation we should add a parenthesis around the two terms:

@@

catmath(x^(y + z))
```

As we can see, R\ automatically put `y + z` in parentheses even if, from a mathematical point of view, the brackets are not useful. To solve this issue two solutions are possible, the first one is cumbersome since it requires many specifications. The second, the shortcut, uses a syntax similar to LateX and puts the power in curly brackets. In both cases the output produced is consistent with the mathematical notation.

```{r, results='asis', echo=TRUE}
# without parenthesis
mathout(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))

@@

blabla `quote(a * (b + c))[[3]][[1]]`
## Absolute value

@@
